### 2018-03-15 paper/1-intro.tex
Lots more from Simon

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.  Despite this, the challenges of scale have driven
large software firms like Microsoft, Facebook, and Google to develop
their own build systems, each with its own choices and idiosyncrasies.

Seldom do people ask questions like ``What does it mean for my build
system to be correct?'' or ``What are the trade-offs between different
approaches?''.  Complex build systems use subtle algorithms, but they
are often hidden away, and not the object of study.

In this paper we offer a general framework in which to discuss these
questions in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code).  Specifically we make
these contributions:
\begin{itemize}
\item We describe some simple but novel abstractions that
crisply encapsulate what a build system is.
\item We show that we can instantiate
these abstractions to describe the essence of a variety of different
build systems, including \Make, \Shake, \Bazel, and \Excel, each in
a dozen lines of code or so.
Doing this in a single setting allows
the differences and similarities between these huge systems to be
brought out clearly.
\item Build systems vary on many axes,
including: static vs dynamic dependencies; cloud-build, including
shallow vs deep; deterministic vs non-deterministic build rules;
early cut-off; self-tracking build systems; and persistent metadata.
These properties (which we define in~\S\ref{sec-background}) are often
deeply built into any particular build system.
In contrast, our framework allows them to be distinguished,
reasoned about, and varied.
\item We show that all build system we look at can be broken along two axes - the order in which they build dependencies, and whether they decide to build a particular dependency. Interestingly, these two aspects can be recombined without restriction to produce new build systems.
\item Two particularly desirable properties are dynamic dependencies
and cloud-build; yet no currently-available build system supports
both.  Our framework makes it rather easy to do so, for the first
time.
\end{itemize}
In short, instead of seeing build systems as unrelated
points in space, we can re-envisage them as locations in a landscape,
leading to better understanding of what they do and how they compare,
and suggesting exploration of other (as yet unoccupied points) in the
landscape.

>>>>

Build systems (such as \Make) are big, complicated, and used by every
software developer on the planet.  But they are a sadly unloved part
of the software ecosystem, very much a means to an end, and seldom the
focus of attention.
Seldom do people ask questions like ``What does it mean for my build
system to be correct?'' or ``What are the trade-offs between different
approaches?''.  Moreover, complex build systems use subtle algorithms, but they
are often hidden away, and not the object of study.

Recently, the challenges of scale have driven large software firms
like Microsoft, Facebook, and Google to develop their own build
systems, each with its own choices and idiosyncrasies.
In this paper we offer a general framework in which to understand
and compare build systems,
in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code).  Specifically we make
these contributions:
\begin{itemize}
\item Build systems vary on many axes,
including: static vs dynamic dependencies; cloud-build, including
shallow vs deep; deterministic vs non-deterministic build rules;
early cut-off; self-tracking build systems; and persistent metadata.
In \S\ref{sec-background} we identify some key properties, in the context of
four carefully-chosen build systems.
\item We describe some simple but novel abstractions that
  crisply encapsulate what a build system is (\S\ref{sec-abstractions}),
  allowing us, for example, to speak about what it means for a build system to be correct.

\item We identify two key design choices
  that are typically deeply built into any build system:
  \emph{the order in which dependencies are built} (\S\ref{sec-dependency-orderings})
  and \emph{whether or not a dependency is (re-)built} (\S\ref{sec-out-of-date}).
  These choices turn out to be orthogonal, which leads us to a new
  classification of the design space (\S\ref{sec-design-space}).

\item We show that we can instantiate
our abstractions to describe the essence of a variety of different
real-life build systems, including \Make, \Shake, \Bazel, and \Excel, each in
a dozen lines of code or so (\S\ref{sec-implementations}).
Doing this in a single setting allows
the differences and similarities between these huge systems to be
brought out clearly.

\item Moreover, we can readily mix the ingredients differently to produce the first
  build system that supports both \emph{dynamic dependencies}
  and \emph{cloud build} (\S\ref{sec-implementation-cloud-shake}).

\end{itemize}
In short, instead of seeing build systems as unrelated
points in space, we now see them as locations in a landscape,
leading to better understanding of what they do and how they compare,
and suggesting exploration of other (as yet unoccupied points) in the
landscape.
We discuss engineering aspects in \S\ref{sec-engineering}, and related
work in \S\ref{sec-related}.

### 2018-03-15 paper/3-abstractions.tex
Lots more from Simon

We start by establishing a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets.

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash
initialise :: Hashable v => i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- State monad
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{-2mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}

@@

In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.
Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

Some values must be provided by the user as \emph{input}. For example,
\cmd{main.c} can be edited by the user who relies on the build system to
compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
but are produced in the process of turning inputs into outputs.

We use a cryptographic \emph{hash function}
\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash} for
efficient tracking and sharing of build results, where \hs{Hash} is an abstract
data type equipped with the equality test.

A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
values. It is convenient to assume that a store is total, i.e. it contains a
value for every possible key. We therefore also assume that the type of values
is capable of encoding values corresponding to non-existent keys (missing files
or empty cells). In addition to usual \hs{getValue} and \hs{putValue} operations,
some stores support a dedicated \hs{getHash} operation, which is implemented
more efficiently than by hashing the result of \hs{getValue}\footnote{For
example, Git Virtual File System~\cite{gvfs} stores file hashes and downloads
file contents only on demand.}. The store also holds a value of type \hs{i},
representing the persistent build information maintained by the build
system~--~its ``memory''. Given an \hs{i} and a key-value map \store, we can
\hs{initialise} a store.

@@

Our first main idea is the task-description abstraction:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _     _    = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
\hs{fetch} function and a key. It pattern-matches on the key to see if it has a
task description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
it computes the value of the formula, using \hs{fetch} to find the value of any
keys on which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures are given in Fig.~\ref{fig-types}. We will
return to the need for this generality in~\S\ref{sec-deps}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: (Eq k, Hashable v) => Build Monad () k v -- Drop Hashable? It's confusing here
busy task key store = execState (compute key) store
  where
    compute k = case task compute k of
        Nothing  -> do { s <- get; return (getValue s k) }
        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
\end{minted}
\vspace{1mm}

\noindent
The \hs{busy} build system uses a function
\hs{compute}~\hs{::}~\hs{k}~\hs{->}~\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)}~\hs{v},
which given a key brings the associated value up-to-date and returns it. To do
so, \hs{compute} calls \hs{task} passing itself as an argument, so that \hs{task}
can use \hs{compute} to recursively find the values of its dependencies. If
\hs{task} returns \hs{Nothing} (this key is an input), \hs{compute} simply reads
the result from the store~\hs{s}, which is available in the \hs{State} monad (in
which \hs{compute} runs) via the \hs{get} method~--~see the \hs{State} monad
interface in Fig.~\ref{fig-types}. Otherwise it runs the action \hs{act}
returned by the \hs{task} to produce a resulting value~\hs{v}, stores it
(modifying the store \hs{s}), and returns it.

Given an acyclic task description, the \hs{busy} build system terminates with a
correct result, but it is not a minimal build system (c.f. the
Definition~\ref{def-minimal}). Since \hs{busy} has no memory (\hs{i = ()}), it
cannot keep track of keys it has already built, and will therefore busily
recompute the same keys again and again if they have multiple dependents. We
will develop much more efficient build systems in~\S\ref{sec-implementations}.

Despite limitations, \hs{busy} can easily handle the example \hs{sprsh1}
from the previous subsection~\S\ref{sec-task}. In the GHCi session below we
initialise the store with \cmd{A1} set to 10 and all other cells set to 20.

@@

An alert reader might notice that \hs{busy} accepts \emph{monadic tasks} of
type \hs{Task}~\hs{Monad}, whereas \hs{sprsh1} is a \hs{Task}~\hs{Applicative}.
This indicates that \hs{busy} can cope with tasks containing dynamic
dependencies, which we discuss in the next subsection~\S\ref{sec-task-monad}.

@@

As explained in~\S\ref{sec-background-excel}, some build tasks have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such build tasks correspond to the type
\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: IF(C1=1,A1,A2)      C1: 1
A2: 20
\end{minted}

\noindent
The task corresponding to the cell \cmd{B1} statically depends on \cmd{C1} and
dynamically depends on either \cmd{A1} or \cmd{A2}. We can express this using
our task abstraction as follows\footnote{The spreadsheet example that uses
the \hs{INDIRECT} function can be expressed very similarly: simply replace the
line containing the \cmd{if} statement with \hs{fetch ("A" ++ show c1)}.}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh2 :: Task Monad String Integer
sprsh2 fetch "B1" = Just $ do c1 <- fetch "C1"
                              if c1 == 1 then fetch "A1" else fetch "A2"
sprsh2 _     _    = Nothing
\end{minted}
\vspace{1mm}

\noindent
The main difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
\hs{fetch} different keys depending on whether \hs{c1}~\hs{==}~\hs{1} or not.

Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
rebuilds every dependency it encounters, it is easy for it to handle dynamic
dependencies. For minimal build systems, however, dynamic dependencies and hence
monadic tasks are much more challenging, as we will see
in~\S\ref{sec-implementations}.

@@

We can now say what it means for a build system to be \emph{correct}, something
that is seldom stated formally. Before proceeding, we need to introduce a
useful utility function called \hs{execute} that executes a \hs{task} on a
\hs{key} by reading its dependencies from a supplied \hs{store}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Identity a = Identity { runIdentity :: a }
\end{minted}
\vspace{0mm}
\begin{minted}[xleftmargin=10pt]{haskell}
execute :: Task Monad k v -> Store i k v -> k -> Maybe v
execute task store key = runIdentity <$> task (Identity . getValue store) key
\end{minted}
\vspace{1mm}

\noindent
This function is simpler than it might look: it fixes the `impedance mismatch'
between the pure function \hs{getValue}~\hs{store}, whose type is \store, and
the \hs{fetch} argument of the \hs{task}, whose type must be
\hs{k}~\hs{->}~\hs{f}~\hs{v} for some \hs{f}. To fix the mismatch, we wrap the
result of the pure function in the \hs{Identity} monad: the function
\hs{Identity}~\hs{.}~\hs{getValue}~\hs{store} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now need to get rid
of the \hs{Identity} wrapper by applying \hs{runIdentity} to the contents of
\hs{Maybe}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
build  :: Build c i k v
task   :: Task c k v
key    :: k
store  :: Store i k v
result :: Store i k v
result = build task key store
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of executing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{execute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
        \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{result}, \hs{store} and \hs{ideal} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies with
    respect to the \hs{result}:
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{store}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    \vspace{-6mm}
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    where the implementation of \hs{inputs} will be provided in the next
    subsection~\S\ref{sec-deps}, which is dedicated to the task dependency
    analysis. This requirement guarantees that (i) no inputs were corrupted
    during the build, and (ii)~the \hs{ideal} store has constraints not only
    on the output, but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}

@@

To extract the dependencies of monadic tasks we introduce the function \hs{track}:
a combination of \hs{execute} and \hs{dependencies} that takes a store as input
and computes both the resulting value and the list of its dependencies:

>>>>

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash
initialise :: Hashable v => i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- State monad
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{-2mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems.
% that allows
% us to abstract away from specific application domains, such as software
% development or spreadsheets.
Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

@@

\paragraph{Keys and values.} In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.

\paragraph{Input, output, and intermediate values}
Some values must be provided by the user as \emph{input}. For example,
\cmd{main.c} can be edited by the user who relies on the build system to
compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
but are produced in the process of turning inputs into outputs.

\paragraph{The store.}
A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
values. It is convenient to assume that a store is total, i.e. it contains a
value for every possible key. We therefore also assume that the type of values
is capable of encoding values corresponding to non-existent keys (missing files
or empty cells).

\paragraph{Hashes.}
We use a cryptographic \emph{hash function}
\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash} for
efficient tracking and sharing of build results, where \hs{Hash} is an abstract
data type equipped with the equality test.
In addition to usual \hs{getValue} and \hs{putValue} operations,
some stores support a dedicated \hs{getHash} operation, which is implemented
more efficiently than by hashing the result of \hs{getValue}\footnote{For
example, Git Virtual File System~\cite{gvfs} stores file hashes and downloads
file contents only on demand.}.

\paragraph{Persistent build information.}
The store also holds a value of type \hs{i},
representing the persistent build information maintained by the build
system~--~its ``memory''.

% Given an \hs{i} and a key-value map \store, we can
% \hs{initialise} a store.

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _     _    = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
\hs{fetch} function and a key. It pattern-matches on the key to see if it has a
task description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
it computes the value of the formula, using \hs{fetch} to find the value of any
keys on which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures are given in Fig.~\ref{fig-types}.   We will
explain why in the next subsection (\S\ref{sec-general-build}).

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: (Eq k, Hashable v) => Build Monad () k v -- Drop Hashable? It's confusing here
busy task key store = execState (compute key) store
  where
    compute :: k -> State (Store i k v) v
    compute k = case task compute k of
        Nothing  -> do { s <- get; return (getValue s k) }
        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
\end{minted}
\vspace{1mm}

\noindent
The \hs{busy} build system uses a helper function
\hs{compute} that, given a key, brings the key up to date in the store, and returns its value.
The function \hs{compute} runs in the standard Haskell \hs{State} monad -- see
Figure~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
To bring a key up to date, \hs{compute} asks the task description \hs{task} how
to compute \hs{k}.
If \hs{task} returns \hs{Nothing} the key is an input, so \hs{compute} simply reads
the result from the store.
Otherwise \hs{compute} runs the action \hs{act}
returned by the \hs{task} to produce a resulting value~\hs{v}, records
the new key/value mapping in the store, and returns \hs{v}.
Notice that \hs{compute} passes itself to \hs{task} as an argument,
so that the latter can use \hs{compute} to recursively find the
values of \hs{k}'s dependencies.

Now we can see why the \hs{Task} abstraction needed to be abstracted over \hs{f}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
The \hs{busy} build system instantiates \hs{c} to \hs{Monad} and \hs{f} to \hs{State (Store i k v)},
so that the call-back to \hs{fetch :: k -> f v} can side-effect the \hs{Store},
thereby allowing successive calls to \hs{fetch} to communicate with one another.
We will see why we want \hs{Task} to be \emph{polymorphic} in \hs{f} in \S\ref{sec-deps}.

Given an acyclic task description, the \hs{busy} build system terminates with a
correct result, but it is not a \emph{minimal} build system
(Definition~\ref{def-minimal}). Since \hs{busy} has no memory (\hs{i=()}), it
cannot keep track of keys it has already built, and will therefore busily
recompute the same keys again and again if they have multiple dependents. We
will develop much more efficient build systems in~\S\ref{sec-implementations}.

Nevertheless, \hs{busy} can easily handle the example \hs{sprsh1}
from the previous subsection~\S\ref{sec-task}. In the GHCi session below we
initialise the store with \cmd{A1} set to 10 and all other cells set to 20.

@@

An alert reader might notice that \hs{busy} accepts \emph{monadic tasks} of
type \hs{Task}~\hs{Monad}, whereas \hs{sprsh1} is a \hs{Task}~\hs{Applicative}.
This indicates that \hs{busy} can cope with tasks containing dynamic
dependencies, which we discuss next (\S\ref{sec-task-monad}).

@@

As explained in~\S\ref{sec-background-excel}, some task descriptions have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such task descriptions correspond to the type
\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Recall the following spreadsheet
example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: IF(C1=1,B2,A2)      C1: 1
A2: 20      B2: IF(C1=1,A1,B1)
\end{minted}

\noindent
Here, \cmd{B1} and \cmd{B2} statically depend on each other, but Excel (which
uses dynamic dependencies) is perfectly happy. We can express this using
our task abstraction as follows\footnote{The spreadsheet example that uses
the \hs{INDIRECT} function can be expressed very similarly: simply replace the
line containing the \cmd{if} statement with \hs{fetch ("A" ++ show c1)}.}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh2 :: Task Monad String Integer
sprsh2 fetch "B1" = Just $ do c1 <- fetch "C1"
                              if c1 == 1 then fetch "B2" else fetch "A2"
sprsh2 fetch "B2" = Just $ do c1 <- fetch "C1"
                              if c1 == 1 then fetch "A1" else fetch "B1"
sprsh2 _     _    = Nothing
\end{minted}
\vspace{1mm}

\noindent
The main difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
then \hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.

Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
rebuilds every dependency it encounters, it is easy for it to handle dynamic
dependencies. For minimal build systems, however, dynamic dependencies, and hence
monadic tasks, are much more challenging, as we will see
in~\S\ref{sec-implementations}.

@@

We can now say what it means for a build system to be \emph{correct}, something
that is seldom stated formally. Our intuition is this: \emph{when the build system
  returns, the specified key, and all its dependencies, should be up to date}.
What does ``up to date'' mean?  It means that if we recompute the value of the
key (using the task description, and the final store), we should get exactly
the same value as we see in the final store.

To express this formally we need an auxiliary
function \hs{recompute}, that computes the value of a key in a given store
\emph{without attempting to update any dependencies}:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Identity a = Identity { runIdentity :: a }
\end{minted}
\vspace{0mm}
\begin{minted}[xleftmargin=10pt]{haskell}
recompute :: Task Monad k v -> Store i k v -> k -> Maybe v
recompute task store key = runIdentity <$> task (Identity . getValue store) key
\end{minted}
\vspace{1mm}
\noindent
\todo{SLPJ}{Why monad?}
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad.
The use of \hs{Identity} just fixes the `impedance mismatch'
between the pure function \hs{getValue}~\hs{store}, whose type is \store, and
the \hs{fetch} argument of the \hs{task}, whose type must be
\hs{k}~\hs{->}~\hs{f}~\hs{v} for some \hs{f}. To fix the mismatch, we wrap the
result of the pure function in the \hs{Identity} monad: the function
\hs{Identity}~\hs{.}~\hs{getValue}~\hs{store} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now need to get rid
of the \hs{Identity} wrapper by applying \hs{runIdentity} to the contents of
\hs{Maybe}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
task          :: Task c k v
key           :: k
store, result :: Store i k v
result = build task key store
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of executing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{recompute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
        \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{result}, \hs{store} and \hs{ideal} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies with
    respect to the \hs{result}:
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{store}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    \vspace{-6mm}
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    where the implementation of \hs{inputs} will be provided in the next
    subsection~\S\ref{sec-deps}, which is dedicated to the task dependency
    analysis. This requirement guarantees that (i) no inputs were corrupted
    during the build, and (ii)~the \hs{ideal} store has constraints not only
    on the output, but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}

@@

To extract the dependencies of monadic tasks we introduce the function \hs{track}:
a combination of \hs{recompute} and \hs{dependencies} that takes a store as input
and computes both the resulting value and the list of its dependencies:
